Adapting Duplicate Filters to a Sliding Window Context

### 2020-03-10 qhtv2.tex
wip

\paragraph{Duplicate detection problem.}
We consider the following problem (duplicate detection problem, DDP): given a stream $E_n = \{e_1, e_2, \dotsc, e_n\}$ and a \enquote{new} item $e^\star$, find whether $e^\star \in E_n$. It is understood that the stream $E_n$ is updated at every time increment. An \emph{efficient} solution to this problem provides an answer with bounded error $\epsilon$, \todo[A verifier using polynomially many resources (time and memory) in $1/\epsilon$].

@@

This is however out of the scope of our work, as we consider an element to be a duplicate if and only if the exact same element has already appeared in the stream. \emph{Approximate} detection states that no attempt will be done to reach perfect accuracy, but rather to limit as much as possible the errors. One can prove \cite[Theorem 2.1]{10.1145/3297280.3297335} that perfect detection of duplicates in a stream with $U$ possible values requires $U$ bits of memory: even for modest situations (e.g. 64-bit nonces, corresponding to $2^{64}$ possible values) this is quickly impractical.

@@

\begin{definition}[Probability of false positive, of false negative]
	Let $E = \{e_1, \dotsc\}$ be a stream, $\mathcal F$ be a filter. The \emph{probability of false positive} (resp. \emph{probability of false negative}) of $\mathcal F$ after $i$ elements, denoted $\FP_i$ (resp. $\FN_i$), is defined as the proabability that $e_{i+1}$ is a false positive (resp. false negative).
\end{definition}

@@

\subsection{Filter Saturation and Sliding Windows}
Unfortunately, duplicate detection over an infinite stream has been identified as an ill-posed problem. As a matter of fact, as the number of bits of information of the stream grow to infinity, the filter can only remember an ever diminishing proportion of those bits, eventually having no advantage compared to a filter answering randomly.

@@

@@

When $|\Gamma| > 2 \log_2 w$ (which is the case of most practical use cases) this DDF outperforms the naive strategy\footnote{The naïve strategy consisting of storing the $w$ elements of the sliding window, requiring $w |\Gamma|$ bits of memory.}, both in terms of time and memory.

Note that unlike the DDP, we do not necessarily encounter filter saturation for the sliding window DDP, which makes it possible for different DDF to have different performances in the asymptotic regime and thus to be ranked.

Another remark is that when $M \approx w\log_2w$, the error rate is minimal but $O(w)$ operations ares required (although this complexity may be reduced using parallelized implementations). Detecting duplicates faster in this context remains an open question.

@@

\begin{proof}
	Let $E = \{e_1, \dotsc, e_m, \dotsc\}$ be a stream, let $w$ be a sliding window and let $e^* \notwdup E$. We first take into account the remark in \ref{subsub:number_elements} that the filter $\mathcal Q$ may contain more than $w$ elements, and hence potentially contain $e$ in memory, in the last subfilter of the queue. However, such event is quite rare, and can be neglected: let $\delta = cL - w$ the number of additional elements that can be stored in $\mathcal Q$. The probability that $e^*$ has occurred in $\{e_{m-cL}, \dotsc e_{m-w+1}\}$ is $1-\left(1-\frac1{|\Gamma|}\right)^\delta \approx \frac{\delta}{|\Gamma|} \approx 0$ since $\delta$ is likely to be small, while $|\Gamma|$ is too big for being stored in memory.

	Hence, we can neglect the probability that $e^*$ is present in the filter. Therefore, $e^*$ is a false positive if and only if at least one subquery $\mathcal F_i.\mathsf{Lookup(e^*)}$ returns \DUPLICATE. Conversely, $e^*$ is not a false positive if and only if all subqueries $\mathcal F_i.\mathsf{Lookup(e^*)}$ return \UNSEEN. This can be rewritten as $e^*$ is not a false positive if and only if $e^*$ is not a false positive for each subfilter.

@@

\begin{proof}
	Let $E = \{e_1, \dotsc, e_m, \dotsc\}$ be a stream, let $w$ be a sliding window and let $e^* \wdup E$.
	We first observe that if $cL > w$, there might be some duplicates of $e^*$ that happened before the current sliding window, but might still be remembered by the oldest filter. However, as argued in the previous section, the probability of such an event is negligible.

	Hence, we consider that the only duplicates are part of the sliding window. $e^*$ is a false negative if and only if all subfilters $\mathcal F_i$ answer $\mathcal F_i.\mathsf{Detect}(e^*) = \UNSEEN$.

	There can be two cases:
	\begin{itemize}
		\item $e^*$ is present in $\mathcal F_i$'s sub-sliding window 
		\item $e^*$ is present in $\mathcal F_i$'s sub-sliding window

	\end{itemize}

In the first case, $\mathcal F_i.\mathsf{Detect}(e^*)$ returns \UNSEEN if and only if $e^*$ is a false negative for $\mathcal F_i$. Hence, in this case, the probability of $\mathcal F_i$ to return \UNSEEN is $\FN_{\mathcal F,c}$. For the subfilter $\mathcal F_0$, however, the probability is $\FN_{\mathcal F, m\%c}$.

In the second case, $\mathcal F_i.\mathsf{Detect}(e^*)$ returns \UNSEEN if and only if $e^*$ is not a false positive for $\mathcal F_i$. Hence, in this case, the probability of $\mathcal F_i$ to return \UNSEEN is $1-\FP_{\mathcal F,c}$. For the subfilter $\mathcal F_0$, however, the probability is $1-\FP_{\mathcal F, m\%c}$.

Now, the probability $p_c$ for $e^*$ to be in subfilter $\mathcal F_i$ (which has had $c$ insertions) sub-sliding window is

\begin{align*}
	p_c =& Pr[\text{$e^*$ is in }\mathcal F_i \text{ sub-sliding window | } e^* \wdup E]\\
	=& \frac{Pr[\text{$e^*$ is in }\mathcal F_i \text{ sub-sliding window } \cap e^* \wdup E]}{Pr[e^* \wdup E]}\\
	=& \frac{Pr[\text{$e^*$ is in }\mathcal F_i \text{ sub-sliding window}]}{Pr[e^* \wdup E]}\\
	=& \frac{1 - Pr[\text{$e^*$ is not in }\mathcal F_i \text{ sub-sliding window }]}{1 - Pr[e^* \notwdup E]}\\
	p_c=& \frac{1 - \left(1 - \frac 1{|\Gamma|}\right)^c}{1 - \left(1-\frac{1}{|\Gamma|}\right)^w}
\end{align*}

@@

Let us consider a duplicate $e^* \wdup E$. If, in the sliding window, all duplicates only happened at least $(c-\frac12)L$ epochs ago (which happens with probability $\left(1-\frac1{|\Gamma|}\right)^{(c-\frac12)L}$), we know that no duplicates happened in any subfilter's sliding window. By using the same reasoning as the theorem's proof, we get a probability of false negative of around $\left[0 + \left(1-\frac1{|\Gamma|}\right)^{(c-\frac12)L}(1-\FP_{\mathcal F,c})\right]^L$. Given that $cL \approx w$ which is a constant, we get that the probability of false positive mostly varies, in this case, with $\left(1-\FP_{\mathcal F,c}\right)^L$.

@@

\section{Application on Existing DDFs}
As our queuing filter relies on another DDF subfilter, we must pick one existing DDF from the literature. However, testing all DDFs may be tedious. Instead, we propose a ranking on existing DDFs. As \cite{10.1145/3297280.3297335} mentioned, filters all reach a saturation state. The idea is then to order DDFs by their resistance to saturation --i.e., by ranking them by the time they take to reach saturation.

@@

Note: given we are working on selecting the best subfilter for our queueing filter. As such, in this section we are \emph{not} working on a sliding window.

\subsection{Optimal Error Probability on a DDF}
    \begin{theorem}
	Let $E$ be a stream of $n$ elements uniformly selected from an alphabet of size $|\Gamma|$. For any DDF of size $M$, the error probability $ER_n$ (consisting of the sum of the FP and FN probabilities) is such that
	$$ER_n \geq 1 - \frac{1 - \left(1 - \frac{1}{|\Gamma|}\right)^{M}}{1 - \left(1 - \frac 1{|\Gamma|}\right)^n}$$ for any $n > M$.

	Especially, for any filter of size $M$, the asymptotic error rate $ER_\infty$ is such that $ER_\infty \geq \left(1 - \frac{1}{|\Gamma|}\right)^{M}$.
\end{theorem}

@@

	Let us derive the error rate of such a filter.
	If an element is already stored in the filter, then the optimal filter will necessarily answer \DUPLICATE. On the other hand, if the element is not in memory, a perfect filter has no better option than answering randomly. Let $p$ be the probability that a filter answers \DUPLICATE when an element is not in memory.

	An unseen element, by definition, will be unseen in the $M$ last elements of the stream, and hence will not be in the filter's memory, the filter will return \UNSEEN with probability $1-p$. For this reason, this optimal filter has an FP probability of $p$.

	On the other hand, a duplicate element $e^* \dup E$ is classified as \UNSEEN if and only if it was not seen in the last $M$ elements of the stream, and the filter answers \UNSEEN. Let $D$ be the event "\textsf{There is at least one duplicate in the stream} and $C$ be the event \textsf{There is a duplicate of $e^*$ in the $M$ previous elements of the stream}.

	We have that $e^*$ triggers a false negative with probability

	\begin{align*}
	P_n &= (1 - P[C | D])(1-p) =  \left(1 - \frac{P[C \cap D]}{P[D]}\right)(1-p)\\
	& = \left(1 - \frac{P[C]}{P[D]}\right)(1-p) =  \left(1 - \frac{1 - P[\bar C]}{1 - P[\bar D]}\right)(1-p) \\
	P_n &=  \left(1 - \frac{1 - \left(1 - \frac 1U\right)^M}{1 - \left(1 - \frac 1U\right)^n}\right)(1-p)
	\end{align*}

	Hence, the error rate of the perfect filter is $ER_n = P_n + p =  \left(1 - \frac{1 - \left(1 - \frac{1}{U}\right)^{M}}{1 - \left(1 - \frac 1U\right)^n}\right)(1-p) + p =$ $ 1 - \frac{1-\left(1 - \frac{1}{U}\right)^{M}}{1 - \left(1 - \frac 1U\right)^n}(1-p)$, which is minimized when $p = 0$.

@@

However, one can observe that a small $L$ is undesirable, as was exposed in \ref{sub:fnr}, because it would lead to a greater number of false negatives.

\todo[observation experimentale ?]

@@

\subsection{Comparison of QHT and QHT-Queueing Filter}
In this section, we compare

>>>>

\paragraph{Duplicate detection problem.}
We consider the following problem (duplicate detection problem, DDP): given a stream $E_n = \{e_1, e_2, \dotsc, e_n\}$ and a \enquote{new} item $e^\star$, find whether $e^\star \in E_n$, using only $M$ bits of memory. It is understood that the stream $E_n$ is updated at every time increment. An \emph{efficient} solution to this problem provides an answer with bounded error.

@@

This is however out of the scope of our work, as we consider an element to be a duplicate if and only if the exact same element has already appeared in the stream. \emph{Approximate} detection states that no attempt will be done to reach perfect accuracy, but rather to limit as much as possible the errors. As a matter of fact, in the context of approximate DDT, it is assumed that $M$ is to small for classical naïve solutions. One can prove \cite[Theorem 2.1]{10.1145/3297280.3297335} that perfect detection of duplicates in a stream with $U$ possible values requires $U$ bits of memory: even for modest situations (e.g. 64-bit nonces, corresponding to $2^{64}$ possible values) this is quickly impractical.

@@

\begin{definition}[Probability of false positive, of false negative]
	Let $E = \{e_1, \dotsc\}$ be a stream, $\mathcal F$ be a filter. The \emph{probability of false positive} (resp. \emph{probability of false negative}) of $\mathcal F$ after $i$ elements, denoted $\FP_i$ (resp. $\FN_i$), is defined as the probability that $e_{i+1}$ is a false positive (resp. false negative).
\end{definition}

@@

\subsection{Filter Saturation and Sliding Windows}
Unfortunately, efficient duplicate detection over an infinite stream has been identified as an unsolvable problem \cite{10.1145/3297280.3297335}. As a matter of fact, as the number of bits of information of the stream grow to infinity, the filter can only remember an ever diminishing proportion of those bits, eventually having no advantage compared to a filter answering randomly.

@@

Our goal is to minimise both the false positive and false negative rates of a DDF, given a fixed and limited amount of memory. Before going into details, we first explore the limits of the problem, i.e., when perfect or very performant solutions exist.

\begin{theorem}
	For $M = w (\log_2(w) + 1) \log_2(|\Gamma|)$, the sliding window DDP can be solved with 0 errors in constant time.
\end{theorem}

\begin{proof}
	To store $w$ elements from a sliding window, the naive idea is to use a queue $Q$ of $w$ elements. However, this yields a complexity of $O(w)$ for lookup, which might be impractical. Thus, we create a second structure for the sake of quickly answering lookup questions.

	The dictionary $D$ has the form $\Gamma \Rightarrow \mathbb N$: keys are elements from the alphabet, values are integers serving as counters.

	During insertion of $e$, $e$ is appended to the queue and $D[e]$ is incremented. If the key $e$ did not exist in $D$, it is created first.

	The insertion of an element means that another element, at the end of the sliding window, is removed. For this, we pop the last element of $Q$, and get $e_{last} \gets Q.pop()$. Then, we decrement $D[e_{last}]$ and if $D[e_{last}] = 0$, the key is deleted from $D$.

	Lookup of an element $e^\star$ is simply done by looking whether the key $D[e^\star]$ exists, which is done in constant time.

	It is immediate to see that such a structure has a 0 error rate over the sliding window and that both insertion and lookup times are constant.

	The size of the queue is $w\log_2 |\Gamma|$, the size of the dictionary is $w \log_2 |\Gamma| \log_2 w$ (as the dictionary cannot have more than $w$ keys as the same time, and a counter cannot go over $w$, thus being less than $\log_2 w$ bits long), hence the claim. 
\end{proof}

However, this implementation depends on $\log_2 |\Gamma|$, which might be arbitrarily large. We thus propose another solution, with a very low error rate but constant time.

@@

When $\log_2|\Gamma| > 2 \log_2 w$ (which is the case of most practical use cases) this DDF outperforms the naïve strategy\footnote{The naïve strategy consisting of storing the $w$ elements of the sliding window, requiring $w \log_2|\Gamma|$ bits of memory.}, both in terms of time and memory.

Note that this solution has a time complexity of $O(w)$. Using a dictionary like in the previous proof, we get a filter with an error rate of about $\frac 1w$ and constant time for insertion and lookup, but requiring $5w\log_2 w$ bits of memory.

Hence, very good solutions for situations where $M \approx 5w\log_2 w$ already exist, but efficient solutions and lower bounds on the error rates for smaller amounts of memory are still to be found.

@@

\begin{proof}
	Let $E = \{e_1, \dotsc, e_m, \dotsc\}$ be a stream, let $w$ be a sliding window and let $e^\star \notwdup E$. We first take into account the remark in \ref{subsub:number_elements} that the filter $\mathcal Q$ may contain more than $w$ elements, and hence potentially contain $e$ in memory, in the last subfilter of the queue. However, such event is quite rare, and can be neglected: let $\delta = cL - w$ the number of additional elements that can be stored in $\mathcal Q$. The probability that $e^\star$ has occurred in $\{e_{m-cL}, \dotsc e_{m-w+1}\}$ is $1-\left(1-\frac1{|\Gamma|}\right)^\delta \approx \frac{\delta}{|\Gamma|} \approx 0$ since $\delta$ is likely to be small, while $|\Gamma|$ is too big for being stored in memory.

	Hence, we can neglect the probability that $e^\star$ is present in the filter. Therefore, $e^\star$ is a false positive if and only if at least one subquery $\mathcal F_i.\mathsf{Lookup(e^\star)}$ returns \DUPLICATE. Conversely, $e^\star$ is not a false positive if and only if all subqueries $\mathcal F_i.\mathsf{Lookup(e^\star)}$ return \UNSEEN. This can be rewritten as $e^\star$ is not a false positive if and only if $e^\star$ is not a false positive for each subfilter.

@@

\begin{proof}
	Let $E = \{e_1, \dotsc, e_m, \dotsc\}$ be a stream, let $w$ be a sliding window and let $e^\star \wdup E$.
	We first observe that if $cL > w$, there might be some duplicates of $e^\star$ that happened before the current sliding window, but might still be remembered by the oldest filter. However, as argued in the previous section, the probability of such an event is negligible.

	Hence, we consider that the only duplicates are part of the sliding window. $e^\star$ is a false negative if and only if all subfilters $\mathcal F_i$ answer $\mathcal F_i.\mathsf{Detect}(e^\star) = \UNSEEN$.

	There can be two cases:
	\begin{itemize}
		\item $e^\star$ is present in $\mathcal F_i$'s sub-sliding window
		\item $e^\star$ is not present in $\mathcal F_i$'s sub-sliding window
	\end{itemize}

In the first case, $\mathcal F_i.\mathsf{Detect}(e^\star)$ returns \UNSEEN if and only if $e^\star$ is a false negative for $\mathcal F_i$. Hence, in this case, the probability of $\mathcal F_i$ to return \UNSEEN is $\FN_{\mathcal F,c}$. For the subfilter $\mathcal F_0$, however, the probability is $\FN_{\mathcal F, m\%c}$.

In the second case, $\mathcal F_i.\mathsf{Detect}(e^\star)$ returns \UNSEEN if and only if $e^\star$ is not a false positive for $\mathcal F_i$. Hence, in this case, the probability of $\mathcal F_i$ to return \UNSEEN is $1-\FP_{\mathcal F,c}$. For the subfilter $\mathcal F_0$, however, the probability is $1-\FP_{\mathcal F, m\%c}$.

Now, the probability $p_c$ for $e^\star$ to be in subfilter $\mathcal F_i$ (which has had $c$ insertions) sub-sliding window is

\begin{align*}
	p_c =& Pr[\text{$e^\star$ is in }\mathcal F_i \text{ sub-sliding window | } e^\star \wdup E]\\
	=& \frac{Pr[\text{$e^\star$ is in }\mathcal F_i \text{ sub-sliding window } \cap e^\star \wdup E]}{Pr[e^\star \wdup E]}\\
	=& \frac{Pr[\text{$e^\star$ is in }\mathcal F_i \text{ sub-sliding window}]}{Pr[e^\star \wdup E]}\\
	=& \frac{1 - Pr[\text{$e^\star$ is not in }\mathcal F_i \text{ sub-sliding window }]}{1 - Pr[e^\star \notwdup E]}\\
	p_c=& \frac{1 - \left(1 - \frac 1{|\Gamma|}\right)^c}{1 - \left(1-\frac{1}{|\Gamma|}\right)^w}
\end{align*}

@@

Let us consider a duplicate $e^\star \wdup E$. If, in the sliding window, all duplicates only happened at least $(c-\frac12)L$ epochs ago (which happens with probability $\left(1-\frac1{|\Gamma|}\right)^{(c-\frac12)L}$), we know that no duplicates happened in any subfilter's sliding window. By using the same reasoning as the theorem's proof, we get a probability of false negative of around $\left[0 + \left(1-\frac1{|\Gamma|}\right)^{(c-\frac12)L}(1-\FP_{\mathcal F,c})\right]^L$. Given that $cL \approx w$ which is a constant, we get that the probability of false positive mostly varies, in this case, with $\left(1-\FP_{\mathcal F,c}\right)^L$.

@@

\section{Application on Existing DDFs}
As our queuing filter relies on another DDF subfilter, we must pick one existing DDF from the literature.
\subsection{DDF Subfilter Selection Strategy}
As a matter of fact, it is preferable to use only one kind of DDF as subfilter, as proven by the following theorem.

\begin{theorem}
	An optimal queueing filter only uses one kind of subfilter.
\end{theorem}

\begin{proof}
	One could imagine a queueing filter in which subfilters would alternate between, say, subfilter of type $A$ and of type $B$. However, we show with a simple symmetry argument that this kind of queueing filter is sub-optimal.

	Let us consider a queueing filter relying on several kinds of subfilters. Given that all subfilters play the same role, and that their efficiency is independent of the other subfilters present, we know that one less efficient subfilter can be replaced by a more efficient one, directly improving the efficiency of the queueing filter.
\end{proof}

However, making a theoretical analysis of all DDFs error rates may be tedious. Instead, we propose a ranking on existing DDFs. As \cite{10.1145/3297280.3297335} mentioned, filters all reach a saturation state. The idea is then to order DDFs by their resistance to saturation --i.e., by ranking them by the time they take to reach saturation.

@@

Note: given we are working on selecting the best subfilter for our queueing filter, in this section we are \emph{not} working over a sliding window.

\subsection{Optimal Error Probability on a DDF}\label{sub:optimal_ddf}
    \begin{theorem}
	Let $E$ be a stream of $n$ elements uniformly selected from an alphabet of size $|\Gamma|$. For any DDF of size $M$, the error probability $ER_n$ (consisting of the sum of the FP and FN probabilities) is such that
	$$ER_n \geq 1 - \frac{1 - \left(1 - \frac{1}{|\Gamma|}\right)^{M}}{1 - \left(1 - \frac 1{|\Gamma|}\right)^n}$$ for any $n > M$.

	Especially, for any filter of size $M$, the asymptotic error rate $ER_\infty$ is such that $ER_\infty \geq \left(1 - \frac{1}{|\Gamma|}\right)^{M}$.
\end{theorem}
We remind that in this section, we are working on DDP over the whole stream and not over a sliding window.

@@

	Let us derive the error rate of such a filter.
	If an element is already stored in the filter, then the optimal filter will necessarily answer \DUPLICATE. On the other hand, if the element is not in memory, a perfect filter can choose to answer randomly. Let $p$ be the probability that a filter answers \DUPLICATE when an element is not in memory. An optimal filter will lower the error rate of any filter using the same strategy with a different probability.

	An unseen element, by definition, will be unseen in the $M$ last elements of the stream, and hence will not be in the filter's memory, so the filter will return \UNSEEN with probability $1-p$. For this reason, this filter has an FP probability of $p$.

	On the other hand, a duplicate element $e^\star \dup E$ is classified as \UNSEEN if and only if it was not seen in the last $M$ elements of the stream, and the filter answers \UNSEEN. Let $D$ be the event "\textsf{There is at least one duplicate in the stream} and $C$ be the event \textsf{There is a duplicate of $e^\star$ in the $M$ previous elements of the stream}.

	We have that $e^\star$ triggers a false negative with probability

	\begin{align*}
	\FN_n &= (1 - P[C | D])(1-p) =  \left(1 - \frac{P[C \cap D]}{P[D]}\right)(1-p)\\
	& = \left(1 - \frac{P[C]}{P[D]}\right)(1-p) =  \left(1 - \frac{1 - P[\bar C]}{1 - P[\bar D]}\right)(1-p) \\
	\FN_n &=  \left(1 - \frac{1 - \left(1 - \frac 1{|\Gamma|}\right)^M}{1 - \left(1 - \frac 1{|\Gamma|}\right)^n}\right)(1-p)
	\end{align*}

	Hence, the error probability of the filter is $ER_n = \FN_n + p =  \left(1 - \frac{1 - \left(1 - \frac{1}{|\Gamma|}\right)^{M}}{1 - \left(1 - \frac 1{|\Gamma|}\right)^n}\right)(1-p) + p =$ $ 1 - \frac{1-\left(1 - \frac{1}{|\Gamma|}\right)^{M}}{1 - \left(1 - \frac 1{|\Gamma|}\right)^n}(1-p)$, which is minimized when $p = 0$.

@@

We have observed in \ref{sub:fnr} that the FN is positively correlated with
$\left(1-\FP_{\mathcal F,c}\right)^L$, however the FP is negatively correlated with the same value. As such, decreasing the probability of false positive leads to an increase in the probability of false negatives.

However, practical tests show (see \ref{sub:L_bench}) that setting $L = \sqrt w$ leads to optimal results.

Hence, the (heuristic) optimal solution is to set $L = \sqrt w$.

@@

\subsection{Benchmarking Various Queuing Filters}
In this section, we experimentally prove the heuristic used in \ref{sub:optimal_ddf}, that the best DDF operating on a full stream leads to the best queueing filter operating on a sliding window.

\todo

\subsection{Benchmark of the influence of $L$ parameter}\label{sub:L_bench}

\subsection{Comparison of QHT and QHT-Queueing Filter}
In this section, we compare the efficiency of QHT and QHT-queueing filter. For this, we use both an artificial dataset, as well as a real one, on various sliding window sizes.
